Statistical paradoxes in coronavirus case-counts: selection bias, measurement error, and the COVID-19 pandemic

### 2021-07-12 write-up/covid-writeup.tex
Update covid-writeup.tex
Edits through 4.1

%%  MADE IT TO HERE

\subsection{Estimation of effective reproduction number}
\label{section:r0-estimation}
Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.  Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \citep{Bettencourt2008}.  In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.  Using this relation, a moment-based estimator is given by
$$
R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
$$
Of course, we do not observe $K_t$ and $K_{t-1}$.  Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.  Unfortunately the number of susceptible individuals on day $t$ is unknown.  Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$. We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
$$
\begin{aligned}
\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
\end{aligned}
$$
This implies a similar trade-off as before but on the logarithmic scale.  The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence $\bar Y_t$ but on the fraction of new cases out of those susceptible $\bar K_t$. This leads to differences in when the bias is most pronounced. The final term is the error due to varying numbers of susceptible individuals.  Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the fraction of the population that is susceptible remains large.

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
 \caption{Ratio estimator}
 \label{fig:ratio-bias}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
 \caption{Effective reproductive rate estimator}
 \label{fig:r0-bias}
\end{subfigure}
\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.  Here we assume $f = 0.02$, $FP = 0.005$, $FN = 0.172$, and use a range of relative sampling fraction $M = f_1/f_0$.}
\label{fig:rates}
\end{figure}

@@

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/sir.png}
 \caption{Fraction of new cases in population}
 \label{fig:fracpop}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/sir_rt_comparison.png}
 \caption{Effective reproduction rate estimators}
 \label{fig:eff}
\end{subfigure}
\caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions. Right: comparison of $\hat R_t$ across time with $FN = 0.172$, $FP = 0.005$, and $M = 4$.}
\label{fig:comparison}
\end{figure}

@@

To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.  For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now. Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$. Maximum likelihood estimation follows by maximizing
\begin{equation}
\label{eq:propensity}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
\end{equation}
The first sum only involves individuals observed in the nonprobability sample.
The second sum is over the entire population.  Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.  Typically, this is not possible.  Here we present a method that uses auxiliary information obtained from probability samples as well as probability subsamples from the self-selected population.

>>>>

\subsection{Estimation of effective reproduction number}
\label{section:r0-estimation}
Many epidemiologists argue that tracking the effective reproduction number $R_t$ is the only way to manage through the crisis~\citep{Gabriel2020}.  Under a Poisson likelihood, a simple relation between the trajectory of new cases and the effective reproduction number can be derived \citep{Bettencourt2008}.  In particular, under an SIR model the number of case counts on day $t$, denoted $K_t$, is Poisson distributed with rate $K_{t-1} \exp \left( \gamma (R_t - 1) \right)$ where $K_{t-1} = Y_{t-1}-Y_{t-2}$ is the number of new cases on day $t-1$ and $\gamma$ is the serial interval, which is approximately $7$ days for COVID-19~\citep{Sanche2020}.  Using this relation, a moment-based estimator is given by
$$
R_t = 1 + \frac{1}{\gamma} \log \left( \frac{K_t}{K_{t-1}} \right).
$$
Of course, we do not observe $K_t$ and $K_{t-1}$.  Under SRS of new cases among those susceptible on day $t$, the natural estimator is $S_t \tilde y_t$.  Unfortunately the number of susceptible individuals on day $t$ is unknown.  Here, we study the estimator $\hat R_t = 1 + \frac{1}{\gamma} \log \left( \tilde y_t / \tilde y_{t-1} \right)$. We can again express the statistical error of $\hat R_t - R_t$ in useful terms as follows
$$
\begin{aligned}
\frac{1}{\gamma}\log &\bigg( 1 + \bigg[ \rho_{I_t,K_t} D_{M_t} \sqrt{\frac{1-f_t}{f_t}} CV (K_t)  -\rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \bigg] \\
&\times \left[ 1 - \rho_{I_{t-1},K_{t-1}} D_{M_{t-1}} \sqrt{\frac{1-f_{t-1}}{f_{t-1}}} CV (K_{t-1}) \right] \bigg) - \frac{1}{\gamma} \log \left( \frac{S_t}{S_{t-1}} \right).
\end{aligned}
$$
This implies a similar trade-off as before but on the logarithmic scale.  The error is no longer scaled by $\bar Y_t/\bar Y_{t-1}$ but by the serial interval and does not depend on prevalence but on the fraction of new cases out of those susceptible. This leads to differences in when the bias is most pronounced. Figure~\ref{fig:r0-bias} displays the bias as a function of the relative sampling fraction assuming the population fraction that is susceptible remains large.

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/sir_ratio.png}
 \caption{Ratio estimator}
 \label{fig:ratio-bias}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/sir_rt.png}
 \caption{Effective reproductive rate estimator}
 \label{fig:r0-bias}
\end{subfigure}
\caption{Potential bias due to self-selection and measurement error from a deterministic SIR model with $\beta = 1.2$ (black) and $\gamma = 0.15$.  Here we assume $f = 0.02$, $FP = 0.024$, $FN = 0.30$, and use a range of relative sampling fraction $M = f_1/f_0$.}
\label{fig:rates}
\end{figure}

@@

\begin{figure}
\centering
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/sir.png}
 \caption{Fraction of new cases in population}
 \label{fig:fracpop}
\end{subfigure}%
\begin{subfigure}{.5\textwidth}
 \centering
 \includegraphics[width=.9\linewidth]{../figs/sir_rt_comparison.png}
 \caption{Effective reproduction rate estimators}
 \label{fig:eff}
\end{subfigure}
\caption{Left: fraction infected in two SIR models with $\beta = 1.2$ and $0.9$ respectively and $\gamma = 0.15$ with same initial conditions. Right: comparison of $\hat R_t$ across time with $FN = 0.30$, $FP = 0.024$, and $M = 4$.}
\label{fig:comparison}
\end{figure}

@@

To see this mathematically, let $X_j$ denote a vector of covariates for the $j$th individual in the population, $j=1,\ldots,N$.  For simplicity, the dynamic nature of the outbreak and recoverability of individuals is ignored for now. Suppose the selection indicator $I_j$ is a Bernoulli random variable that depends only on these covariates, i.e., $P(I_j = 1 \mid X_j = x) = \pi (x; \theta)$. Maximum likelihood estimation follows by maximizing
\begin{equation}
\label{eq:propensity}
\sum_{j=1}^N I_j \log \left( \frac{\pi (X_j; \theta)}{1-\pi(X_j; \theta)} \right) + \sum_{j=1}^N \log \left( 1 - \pi (X_j; \theta) \right).
\end{equation}
The first sum only involves individuals observed in the nonprobability sample.
The second sum is over the entire population.  Maximum likelihood estimation therefore requires knowledge of covariate information for every individual in the population.  Typically, this is not possible.  Here we present a method that uses auxiliary information obtained from probability samples.
